---
title: "Class 9"
author: "Christina Liem"
date: "10/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
wisc.df <- read.csv("WisconsinCancer.csv")
```

```{r}
head(wisc.df)
```


```{r}
View(wisc.df)
```

#head(wisc.df)
Here we examine data from XXXX patient samples
How do we do that?

Here we examine data from `r nrow(wisc.df)` patient samples.

how many cases are malignant vs benign?

```{r}
table(wisc.df$diagnosis)
```
In this data set we have 212 cancer and 357 non cancer

```{r}
x <- table(wisc.df$diagnosis)
x
```

In this data set we have `r x["M"]` cancer and `r x["B"]` non cancer

#Hands-on worksheet

Q1. What type of object is returned from the read.csv() function?


Q2. How many observations (i.e. patients) are in this dataset?
569

Q3. How many of the observations have a malignant diagnosis?
table(wisc.df$diagnosis)
212 cases
    
Q4. How many variables/features in the data are suffixed with _mean?
colnames(wisc.df)
10

```{r}
colnames(wisc.df)
```

```{r}
grep("_mean", colnames(wisc.df), value = TRUE)
```

To find out how many there are I can call `length()` on the result of grep
```{r}
length(grep("_mean", colnames(wisc.df), value = TRUE))
```

##Performing PCA

The next step in our analysis is to perform principal component analysis (PCA) on wisc.data
Do we need to scale the data

```{r}
wisc.data <- as.matrix(wisc.df[3:32])
```



# Check column means and standard deviations

```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```



```{r}
round(apply(wisc.data, 2, sd),3)
```
using "2" applies it to the columns
Looks like we need to use `scale=TRUE` here as our data are all over the shop...

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE )
summary(wisc.pr)
```

Why did we use scale = TRUE?
for comparison reasons


Plot PC1 vs PC2 and color by M/B cancer/non-cancer
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2])
```

Add color

```{r}
#with color
plot(wisc.pr$x[, 1:2], col = wisc.df$diagnosis)
```


Q7. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
Look at summary(wisc.pr)
Under cumultative proportion: 0.4427
    

```{r}
x <- summary(wisc.pr)
```

```{r}
x$importance
```

```{r}
x$importance[,"PC1"]
```

The first PC captures `r x$importance[2, "PC1"] *100` of the original vaiance in the data set

Q8. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
Look at cumulative proportion. It takes 3 PCs

```{r}
which(x$importance[3,] > 0.7)[1]
```

Q9. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
7 PCs

```{r}
which(x$importance[3,] > 0.9)[1]
```

#Hierarchial clustering

hclust(x, center =3, nstart = 20)
cutree(hc, k=3)

```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```


```{r}
wisc.hclust <- hclust(data.dist)
```

```{r}
plot(wisc.hclust)
```



```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```


```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```

```{r}
diagnosis <- (wisc.df$diagnosis)
table(wisc.hclust.clusters, diagnosis)
```

The dendrogram above looks awful.

Lets try clustering in PCA space

we will take the results of `prcomp()` and build our distance matrix in PCA space rather than from our raw data.

```{r}
#take the first 7 PCs
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method = "ward.D2")
plot(wisc.pr.hclust)
```


This tree looks much better and we can see that cutting the tree at a height ~70 would yield two clear clusters, lets check


```{r}
plot(wisc.pr.hclust)
abline(h=70, col = "red")
```


And the cluster membership vector can be obtained from `cutree()`

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```

```{r}
table(wisc.df$diagnosis)
```



```{r}
wisc.km <- kmeans(wisc.hclust.clusters, centers= 3, nstart= 20)
table(wisc.km$cluster, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```



```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```


## prediction

We will use the predict() function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

Add these new patients predictions to our main result figure PC1 vs PC2 plot.

```{r}
plot(wisc.pr$x[, 1:2], col = wisc.df$diagnosis)
points(npc[,1], npc[,2], col="blue", pch=15, cex=3)
text(npc[,1], npc[,2], labels=c(1,2), col="white")
```


```{r}
wisc.pr$scale
```



Q22. Which of these new patients should we prioritize for follow up based on your results?
patients clustered in 2 because they have malignant cancer






















