---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
## Setup Bioconductor DESeq2


```{r eval=FALSE}
intsall.packages("BiocManager")
BiocManager::install()
# For this class, you'll also need DESeq2:
BiocManager::install("DESeq2")
```


## Data for today's class
```{r}
counts <- read.csv("airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("airway_metadata.csv", stringsAsFactors = FALSE)

```

Let's have a peak...

```{r}
head(counts)
```


How many genes do we have in this dataset?

```{r}
nrow(counts)
```

How many experiments 
```{r}
ncol(counts)-1
```


Let's make sure metadata id col matches the colnames of counts

```{r}
colnames(counts)[-1]
```

```{r}
metadata$id
```


```{r}
all(colnames(counts)[-1] == metadata$id)
```

The all function is useful for looking accross a vector of logicals

```{r}
all(c(F,T,T))
```



## Analysis: compare the control to drug treated

First we need to access the columns of our countData that are control and then find their mean

```{r}
metadata$dex=="control"
```
```{r}
control.id <- metadata[metadata$dex=="control",]$id
control.id
```

Focus on the counts for control experiments...
```{r}
head(counts[,control.id])
```

Mean counts for control exps
```{r}
control.mean <- rowSums(counts[,control.id])/length(control.id)
names(control.mean) <- counts$ensgene
```

Do this for the drug treated samples

```{r}
treated <- metadata[metadata[,"dex"]=="treated",]
treated.mean <- rowSums( counts[ ,treated$id] )/4 
names(treated.mean) <- counts$ensgene
```

Let's store these together

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

Plot control vs treated

```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```



Good case for a log plot

```{r}
plot(meancounts$control.mean, meancounts$treated.mean, log = "xy")
```




## Fold Change

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

NaN and -Ifn means there's no data for that point

The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero. It turns out that there are a lot of genes with zero expression. Let's filter our data to remove these genes.  Again inspect your results (and the intermediate steps) to see if things make sense.

```{r}
#Test for finding zero entries

x <- c(5,8,0,5)
x == 0
which(x==0)
```

```{r}
x <- data.frame(happy=c(5,6,0,0), sad=c(0,5,5,0))
x==0
```


```{r}
which(x==0, arr.ind=TRUE)
```


Focus on the rows only that have zero entries

```{r}
unique(which(x==0, arr.ind=TRUE)[,1])
```

```{r}
inds <- unique(which(x==0, arr.ind=TRUE)[,1])
x[-inds,]
```

Try it on your own
```{r}
y <- data.frame(happy=c(0,1,2,3,0), sad=c(4,5,6,7,0))
y==0
which(y==0)
```

```{r}
which(y==0, arr.ind=TRUE)
```


```{r}
unique(which(y==0, arr.ind=TRUE)[,1])
```

Relate this technique back to meancounts

```{r}
to.rm <- unique(which(meancounts[,1:2]==0, arr.ind=TRUE)[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```



A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.


```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```


```{r}
sum(up.ind)
sum(down.ind)
```

## Annotations
This is the organism annotation package ("org") for Homo sapiens ("Hs"), organized as an AnnotationDbi database package ("db"), using Entrez Gene IDs ("eg") as primary key. To get a list of all available key types, use:
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

```

See what is available on (org.Hs.eg.db)

```{r}
columns(org.Hs.eg.db)
```


We can use the mapIds() function to add individual columns to our results table. We provide the row names of our results table as a key, and specify that keytype=ENSEMBL. The column argument tells the mapIds() function which information we want, and the multiVals argument tells the function what to do if there are multiple possible values for a single input value. Here we ask to just give us back the first one that occurs in the database.

```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```
```{r}
head(mycounts)
```


## DESeq analysis

```{r}
library(DESeq2)
```

Setup our object for DESeq Analysis


```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```


#Volcano plot

```{r}
plot(res$log2FoldChange, res$padj)
```




```{r}
plot(res$log2FoldChange, -log(res$padj))
```


```{r}
mycols <- rep("gray", length(res$padj))
mycols[abs(res$log2FoldChange) > 2] = "blue"
mycols [(res$padj > 0.05)] = "red"
plot(res$log2FoldChange, -log(res$padj), col=mycols)
```



```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```










