---
title: "Lecture 7"
author: "Christina Liem"
date: "10/23/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Revisit out functions from last day


```{r}
source("http://tinyurl.com/rescale-R")
```


Lets try our rescale() function from the last day

```{r}
rescale(1:10)
```


```{r}
rescale(c(3, 10, NA, 7))
```


What if you add string?


#is.numeric() means is the input for function x is a number
"!" flips TRUE/FALSE

```{r}
is.numeric(5)
```

```{r}
!is.numeric(5)
```

#Should we start like this?
both_na <- function(x,y) {
#something goes here?
}
No! Always start with a simple definiteion of the problem
#Define an example of x and y
Tip: Search for existing functionality to get started

Search internet for R functions that find NA elements

##Write a function for both_na()

```{r}
#from stackoverflow

x <- c( 1, 2, NA, 3, NA) 
y <- c(NA, 3, NA, 3,  4)
is.na(x)
is.na(y)
```



```{r}
which( is.na(x))
```

#which(is.na(x)) tells you which data point is TRUE
working snippet of code
```{r}
is.na(x) & is.na(y)
```

# & means AND 
```{r}
c(T,F,T)
```


```{r}
both_na <- function(x,y) {
  sum(is.na(x) & is.na(y))
}
```


```{r}
both_na(x,y)
```

#you need to use the sum()



```{r}
x <-  c(NA, NA, NA) 
y1 <- c( 1, NA, NA) 
y2 <- c( 1, NA, NA, NA)

both_na(x,y1)

```



```{r}
x <-  c(NA, NA, NA) 
y1 <- c( 1, NA, NA) 
y2 <- c( 1, NA, NA, NA)

both_na(x,y1)


```

```{r}
x <-  c(NA, NA, NA) 
y1 <- c( 1, NA, NA) 
y2 <- c( 1, NA, NA, NA)

both_na(x, y2)

```



```{r}
x <-  c(NA, NA, NA) 
y1 <- c( 1, NA, NA) 
y2 <- c( 1, NA, NA, NA)
y3 <- c(1, NA, NA, NA, NA, NA, NA)

both_na(x, y3)

```


```{r}
plot(1:10, col="red")
```


```{r}
plot(1:10, col=c("red", "blue", "green"))
```


# "==" is something equal
"!=" means not equal

```{r}
both_na <- function(x,y) {
  if(length(x) != length(y)) {
    stop("Input vectors should be the same lenght")
  }
  sum(is.na(x) & is.na(y))
}
```



#refine and polish

both_na3 <- function(x, y) 
{  if(length(x) != length(y)) {    
    stop("Input x and y should be vectors of the same length") 
}   
  na.in.both <- ( is.na(x) & is.na(y) )  
  na.number  <- sum(na.in.both)  
  na.which   <- which(na.in.both)  message("Found ", na.number, " NA's at position(s):",           paste(na.which, collapse=", ") )     return( list(number=na.number, which=na.which) ) }



#How to build a grade() function

```{r}
#student1
s1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 

#student2
s2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
        
```

Steps:
-find the min value
  Search for a function that will help you find the smallest value in a vector


```{r}
#student1
s1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 

#student2
s2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

which.min(s1)

```


-exclude it from the average calculation

```{r}
#student1
s1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 

#student2
s2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

s1[which.min(s1)]

```

```{r}
#student1
s1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 

#student2
s2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

s1[-which.min(s1)]

```

Adding the "-" index will remove the lowest score in this function and call the other scores

```{r}
mean(s1[-which.min(s1)])
```

For student 2


```{r}
mean(s2[-which.min(s2)], na.rm=TRUE)
```

Although this calculates the mean and drops the lowest score, it wouldn't work for a student who was missing the hw
```{r}
s3 <- c(100, NA, NA, NA, NA)
mean(s3[-which.min(s3)], na.rm=TRUE)
```


```{r}
x <- s2
sum(x[-which.min(x)], na.rm=TRUE) #Total minus the worst
```

```{r}
x <- s2
sum(x[-which.min(x)], na.rm=TRUE) #Total minus the worst

#average score
sum(x[-which.min(x)], na.rm=TRUE) / (length(x) - 1)
```

```{r}
grade <- function(x) {
  
sum(x[-which.min(x)], na.rm=TRUE) / (length(x) - 1)
}
```


```{r}
grade(s1)
grade(s2)
grade(s3)
```

#Installing 









