---
title: "Class 11_ Structural Bioinformatics 1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is the PDB database


Download PDB compostition stats from website
http://www.rcsb.org/stats/summary
```{r}
data <- read.csv("Data Export Summary.csv")
data
```

```{r}
total <- sum(data$Total)
total
```



Q1:  Download a CSV file from the PDB site (accessible from “Analyze” -> “PDB Statistics” > “by Experimental Method and Molecular Type”.  Move this CSV file into your RStudio project and determine the percentage of structures solved by X-Ray and Electron Microscopy.  



```{r}
data$Total
```

```{r}
ans <- data$Total/sum(data$Total) * 100
names(ans) <- data$Experimental.Method
round(ans,2)
```


Also can you determine what proportion of structures are protein? Aim to have a rendered GitHub document with working code that yields your answers.


```{r}
round(sum(data$Proteins)/sum(data$Total) * 100, 2)
```

Q2:  Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?


# Working with biomolecular data in R

```{r}
library(bio3d)
pdb <- read.pdb(file = "1hsg")
pbd
```


```{r}
pdb$atom
```


```{r}
pdb$atom[1,"resid"]
```


```{r}
aa321(pdb$atom[,"resid"])
```

```{r}
atom.select.pdb(pdb = pdb, string = "calpha")
```

```{r}
ca.ind <- atom.select(pdb, "calpha")
ca.ind
```

# Print details of the first few selected atoms

```{r}
head( pdb$atom[ca.ind$atom, ] )
```

# And selected xyz coordinates

```{r}
head( pdb$xyz[, ca.ind$xyz] )
```



```{r}
# Select chain A
ca.ind <-atom.select(pdb, chain="A")
# Select C-alphas of chain A
ca.ind <-atom.select(pdb, "calpha", chain="A")
# We can combine multiple selection criteria to return their intersection
cab.ind <-atom.select(pdb, elety=c("CA","CB"), chain="A", resno=10:20)

```



>Q8. Use the Bio3D write.pdb() function to write out a protein only PDB file for viewing in VMD.  Also write out a second separate PDB file for the ligand with residue name MK1

First select "protein" then write out a file: "1hsg_protein.pdb"

```{r}
#atom.select()
atom.select(pdb, "ligand", value = TRUE)
#write.pdb()

```


```{r}
#atom.select()
atom.select(pdb, "protein", value = TRUE)
#write.pdb()

```

```{r}
#atom.select()
prot <- atom.select(pdb, "protein", value = TRUE)
#write.pdb()
write.pdb(prot, file ="1hsg_protein.pdb")

```


Do the same for the ligand


```{r}
lig <- atom.select(pdb, "ligand", value = TRUE)
write.pdb(lig, file = "1hsg_ligand.pdb")
```


```{r}
library(bio3d)
#view lig
view(lig)
view(pdb, "overview", col"sse")
```

"view" doesn't work but it calls up the ligand

```{r}
# The 'devtools' package allows us to install development versions
install.packages("devtools")
```



```{r}
# Load the package
library("bio3d.view")
pdb <-read.pdb("1hel")
# Normal mode analysis calculation
modes <-nma(pdb)
m7 <-mktrj(modes, 
 mode=7,
 file="mode_7.pdb")

view(m7, col=vec2color( rmsf(m7) ))
```








